\documentclass[10pt,letterpaper]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\begin{document} 
%\opening{Dear Unit Owners,} 
 
Recent boards have been doing a great job and I thank them again for all their good work. Having served on boards in the 90's and 00's I know this sort of thing does not happen by magic. Who you elect to represent your interest in your investment matters. I continue to run on a platform of open and honest communication and believe unit owners should be allowed access to what our board is doing. Sunlight is the best disinfectant.

My position is simple: their good work is not a license to break the law. All boards should be aware of the law and adhere to it otherwise they put the building at legal risk. We've been through this once already.  Let's avoid having board members repeat the same mistake.

When I brought this to their attention at a board meeting last year I was met with an unprofessional response. Hopefully we can have a civilized, rational discussion at this year's annual meeting. Don't be surprised if some people get angry and try to prevent this.  Getting angry does not change the facts, but it does discourage decent unit owners from learning the facts about their investment. People that have nothing to hide can give you reasonable answers in a normal tone of voice.

Below is some information about the actions of earlier boards four current members have been hiding from you. Everyone has a right to their private life, but there should be no secret board activities.  

Earlier boards sent me bills for money I did not owe. When I asked how the bills were calculated all I got were legal threats. I hired a lawyer and followed her advice. After two months and over \$1000 in legal fees she told me "These people are not negotiating in good faith, they are out to get you and you need to hire a litigation attorney to protect yourself."

My litigation attorney told me it was the most clear-cut case of selective prosecution he's seen in 13 years of practicing law.  He said we could go to court, and could not guarantee we would win, but if we did and those same people were still on the board they could come after me with more phony things and they could use the building treasury to fund that, while I had to pay out of my own pocket, at least initially.

His advice was to encourage reasonable owners to run for the board and have them look into the matter. As many of you know, I did that and we got an entirely new board that year. But they did not look into it. They spent our money on a lawyer and only settled two weeks before trial.
	
Marty was pretty adamant about not wanting to hear what the ADA both my lawyers suggested I talk to had to say. I think unit owners should be allowed to hear that and would like our next board to look into it. 


%\closing{Keith A. Lewis\\4E\&4F} 
%\cc{Cclist} 
%\ps{adding a postscript} 
%\encl{list of enclosed material} 
\end{document}