\documentclass{article} 
\usepackage{amssymb,amsmath}

\newcommand{\R}{{\bf{R}}}
\newcommand{\onehalf}{\frac{1}{2}}

\begin{document}
In \cite{BlaSch1973}, Fisher Black and Myron Scholes derived a partial
differential equation and boundary conditions that give the 
price of a European option assuming the underlying stock can 
be modeled as geometric Brownian motion and financing is
available from a riskless bond. In \cite{Mer1973} Robert
Merton derives a similar, but different, equation. [(33) in
section 6.]

Their model posits a risk-free bond 
is $X(t) = (R(t), S(t))$ where $dR/R = \rho\,dt$ is a
bond and $dS/S = \mu\,dt + \sigma\,dB$ is a stock with $\rho$, $\mu$,
and $\sigma$ constant and initial conditions $R(0) = 1$, $S(0) = s$.
There are no dividends. They consider portfolios $\Xi(t) = (M(t), N(t))$
consisting of $M$ shares in the bond and $N$ shares in the stock, where
$M$ and $N$ are also Ito diffusions.  The value of the option is $V(t)
= \Xi(t)\cdot X(t) = M(t)R(t) + N(t)S(t)$.

Their {\em self financing} condition is that $d(MR + NS) = M dR + N dS$.
The Ito calculus says $d(MR + NS) = M\,dR + N\,dS + R\,dM + S\,dN + dM\,dR
+ dN\,dS$ so the self financing condition is $0 = R\,dM + S\,dN + dM\,dR +
dN\,dS$.  The financial interpretation is that changing the position $M$
in the bond and $N$ in the stock that were put on at time $t$ to $M +
dM$ and $N + dN$ at time $t + dt$ costs $dM\,(R + dR) + dN\,(S + dS) = 0$.

There exists a function $v\colon [0,\infty)\times\R^2\to\R$ such that
$V(t) = v(t, R(t), S(t))$. Similar functions, $m$ and $n$, exist for
$M$ and $N$.  By the Ito formula applied to $V$ and the self financing
condition, $(v_t + \rho R v_r + \onehalf \sigma^2 S^2 v_{ss}) dt + v_s
dS = M dR + N dS$.  Equating the $dS$ terms gives $N = v_s$. Using $M =
(V - v_s S)/R$, and equating the $dt$ terms yields $v_t + \rho R v_r +
\onehalf\sigma^2 S^2 v_{ss} = \rho(V - v_s S)$, the Black-Scholes/Merton
partial differential equation, almost.

This gives necessary conditions on $M$ and $N$. We show that
they satisfy the self financing condition.

Since $n(t, r, s) = v_s(t, r, s)$,
\begin{equation*}
dN = (v_{st} + \rho R v_{sr} + \onehalf \sigma^2 S^2 v_{sss})\,dt
	+ v_{ss}\,dS.
\end{equation*}
Using $m(t, r, s) = (v - v_s s)/r$, gives
\begin{eqnarray*}
dM &=& \bigl((v_t - v_{st}S)/R + \rho(v_r - v_{sr}S - (v - v_s S)/R)\\
 &&- \onehalf\sigma^2 S^2(v_{ss} + v_{sss}S)/R\bigr)\,dt - (v_{ss}S)/R\,dS.
\end{eqnarray*}

The $dS$ term of $R\,dM + dM\,dR + S\,dN + dN\,dS = R\,dM + S\,dN + dN\,dS$ is 
$-v_{ss}S + v_{ss}S + 0= 0$
and the $dt$ term is
\begin{eqnarray*}
&&\bigl((v_t - v_{st}S) + \rho((v_r - v_{sr}S)R - (v - v_s S)) \\
&&\quad -\onehalf\sigma^2 S^2(v_{ss} + v_{sss}S)\bigr)\\ 
&&+ (v_{st} + \rho R v_{sr} + \onehalf \sigma^2 S^2 v_{sss})S\\
&&+ \sigma^2 S^2 v_{ss}
\end{eqnarray*}

Canceling and combining terms yields
\begin{equation*}
v_t + \rho(v_r R - (v - v_s S))
+\onehalf\sigma^2 S^2v_{ss}
\end{equation*}
Since $v$ satisfies the Black-Scholes/Merton PDE, this
quantity is zero and hence $M$ and $N$ are
self financing.

%This gives necessary conditions on $M$ and $N$. We must show they
%are sufficient. Given $t$, $r$, and $s$, there exists a
%unique strong solution to $(dR, dS) = (\rho R, \mu S) dt
%+ (0, \sigma S)dB$ that satisfies the initial condition 
%$R(t) = r$ and $S(t) = s$ \cite{Oks}.
%For a European option that pays $\phi(S_T)$ at time $T$,
%define $V(t, r, s) = E^(t, r, s)[\phi(S_T)R_T/R_t | \F_t]$.
%The Kolomogorov backward equation \cite{Oks} states
%$V_t = \rho r V_r + \mu_s s V_s + \onehalf \sigma^2 s^2 V_ss$
%and $V(T, r, s) = \phi(s)$. Since $V_r = -V_r/r$ it follows ...


%\bibliographystyle{plain}
%\bibliography{chapter}{}
\end{document}
