\documentclass[11pt,fleqn]{amsproc}
\usepackage{textcomp}

\newcommand{\R}{{\bf R}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\G}{\mathcal{G}}
\renewcommand{\H}{\mathcal{H}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\M}{\mathcal{M}}
\newcommand{\N}{\mathcal{N}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\X}{\mathcal{X}}

\newcommand{\Var}{\mathop{\rm{Var}}}
\newcommand{\Cov}{\mathop{\rm{Cov}}}
\renewcommand{\ker}{\mathop{\rm{ker}}}
\newcommand{\ran}{\mathop{\rm{ran}}}

%\newtheorem{prob}{Problem}[section]
%\newtheorem{cor}[thm]{Corollary}
\newtheorem{xca}{Exercise}

\renewcommand{\labelenumi}{(\alph{enumi})}

\begin{document}

\title{Derivative Securities Final Exam}
%\author{10 May 2010}

\maketitle

For each question, circle the letter corresponding to the answer that
is most correct. Each question is worth 12.5 points.

\begin{xca}
Let $D(t;u,v)$ be the price at time $t$ of a zero coupon bond
over the interval $[u, v]$. Which of the following statements
is always true in every arbitrage free model?
\begin{enumerate}
\item $D(0;1,2) = D(1;1,2)$.
\item $D(0;1,2) < D(1;1,2)$.
\item $D(0;1,2) > D(1;1,2)$.
\item $D(0;1,2) D(1;1,2) = D(0;0,2)$.
\item None of the above.
\end{enumerate}

\end{xca}

\begin{xca}
Let $c(s,k,t)$ be the price of a European call on a non dividend paying
stock with current price $s$, strike $k$, and maturity $t$. Let $C(s,k,t)$
be the price of the corresponding American call option. Assuming positive
interest rates, no arbitrage and perfect liquidity, which of the following
statements are true?

\begin{enumerate}
\item $c(s,k,t) < C(s,k,t)$.
\item $c(s,k,t) \le c(s,k,t')$ when $t \le t'$.
\item $C(s,k,t) \ge \max\{s - k, 0\}$.
\item All of the above.
\item None of the above.
\item Exactly one of (a), (b), (c) is false.
\end{enumerate}

\end{xca}

\begin{xca}
Consider the piecewise linear and continuous function defined by line
segments from $(0,3)$ to $(1,1)$ to $(2,0)$ to $(3,0)$ to $(4,1)$ and
extrapolated linearly. Which of the following statements are true for
portfolios consisting of at most one forward, a cash position, puts and
calls where the puts have strikes less than the those of the calls?

\begin{enumerate}
\item There is a replicating portfolio consisting only of calls and puts that
replicates this payoff.
\item There is a replicating portfolio that has one put having strike 1.
\item There is a replicating portfolio that has one call having strike 1.
\item There is a replicating portfolio that has one put having strike 3.
\item All of the above.
\item Exactly one of (a), (b), (c), (d) is false.
\end{enumerate}
\end{xca}

\begin{xca}
The It\=o integral $\int_0^t e^{\sigma B_s}\,dB_s$ is equal to
\begin{enumerate}
\item $e^{\sigma B_t} - 1 - \int_0^t e^{\sigma B_s}\,ds/2$.
\item $(e^{\sigma B_t} - 1)/\sigma - \sigma\int_0^t e^{\sigma B_s}\,ds$.
\item $(e^{\sigma B_t} - 1)/\sigma - \sigma\int_0^t e^{\sigma B_s}\,ds/2$.
\item $e^{\sigma B_t} - 1 - \sigma\int_0^t e^{\sigma B_s}\,ds/2$.
\item none of the above.
\end{enumerate}
\end{xca}

\begin{xca}
Consider the one period model consisting of a bond, call,
and put where $x = (1, 0.5, 0.5)$,
$X(\omega) = (1, \max\{\omega - 1, 0\}, \max\{2 - \omega, 0\})$,
$0 < \omega < \infty$.
There is an arbitrage consisting of (circle the correct number
in each row)
\begin{enumerate}
\item -1, 0, 1, 2 units of bond
\item -1, 0, 1, 2 units of the call
\item -1, 0, 1, 2 units of the put
\item This model is arbitrage free.
\end{enumerate}
\end{xca}

\begin{xca}
Let $F = fe^{-\sigma^2t/2 + \sigma B_t}$ where $B_t$ is standard
Brownian motion. Let $v = E[\phi(F)]$ where $\phi$ is some
twice differentiable functions. Which of the following are
independent of $\phi$?
\begin{enumerate}
\item $v_{ff}/v_\sigma$.
\item $v_{ff}/v_t$.
\item $v_\sigma/v_t$.
\item All of the above.
\item One of (a), (b), (c) is true.
\item Two of (a), (b), (c) is true.
\end{enumerate}
\end{xca}

\begin{xca}
Let $f = F(0;t,u)$ be today's forward over $[t,u]$, $F = F(t;t,u)$ be
the forward over $[t,u]$ at time $t$, and $D = D_u$ be the stochastic
discount to time $u$. Which of the following statements is true?

\begin{enumerate}
\item $f = E[F] - \Cov(F,D)/E[D]$.
\item $f = E[F] + \Cov(F,D)/E[D]$.
\item $f < E[F]$ if $F$ and $D$ are independent.
\item $f > E[F]$ if $F$ and $D$ are independent.
\item None of the above are true in general.
\end{enumerate}

\end{xca}

\begin{xca}
Consider a European call with strike 50 on a risky zero coupon bond having
100 units notional, one year maturity, and no recovery.  The call expires
one day before bond maturity.  You may suppose the risk-free rate is 3\%
and the hazard rate of default is 2\%. The delta hedge in the risky bond
for this call two days before bond maturity 

\begin{enumerate}
\item is close to zero.
\item is approximately 0.5 shares of the bond.
\item will be close to either 1 or 0 shares of the bond.
\item none of the above.
\item cannot be accurately determined from the given data.
\end{enumerate}
\end{xca}

\end{document}
