\documentclass[11pt,fleqn]{amsproc}
\usepackage{textcomp}

\newcommand{\R}{{\bf R}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\G}{\mathcal{G}}
\renewcommand{\H}{\mathcal{H}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\M}{\mathcal{M}}
\newcommand{\N}{\mathcal{N}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\X}{\mathcal{X}}

\newcommand{\Var}{\mathop{\rm{Var}}}
\newcommand{\Cov}{\mathop{\rm{Cov}}}
\renewcommand{\ker}{\mathop{\rm{ker}}}
\newcommand{\ran}{\mathop{\rm{ran}}}

%\newtheorem{prob}{Problem}[section]
%\newtheorem{cor}[thm]{Corollary}
\newtheorem{xca}{Exercise}

\renewcommand{\labelenumi}{(\alph{enumi})}

\begin{document}

\title{Derivative Securities Final Exam}
\author{11 May 2009}

\maketitle

For each question, circle the letter corresponding to the answer that is most correct. Each question is worth 12.5 points.

\begin{xca}
Consider the payoff $f(x) = -10$, $0 \le x \le 90$; $f(x) = x - 100$, $90 \le x \le 110$; and $f(x) = 10$, $x \ge 110$. In order to produce
this payoff using a cash position, futures contract, and options, it is necessary that
\begin{enumerate}
\item one call with strike 110 be sold,
\item a nonzero cash position be used,
\item a nonzero futures position be used,
\item none of the above,
\item all of the above.
\end{enumerate}
\end{xca}

\begin{xca}
Define a continuous time model $(X_t)_{t\ge0}$ where $X_t = (R_t, S_t)$ with $R_t = \exp(rt)$ and
$S_t = S_0\exp((\mu - \sigma^2/2)t + \sigma B_t)$. Consider the price deflater $\Pi_t = \exp((\nu - \theta^2/2)t
+ \theta B_t)$. The model is arbitrage free
\begin{enumerate}
\item for $\nu = -r$ and any value of $\theta$.
\item for $\nu = -r$ and $\theta = (\mu - r)/\sigma.$
\item only if $\nu = -r$, $\mu = r$, and $\theta = 0$.
\item none of the above.
\end{enumerate}

\end{xca}

\begin{xca}
Consider the one period model $x = (1, 105, 4)$ and $X(\omega) = (1, \omega, (\omega - 100)^+)$ where $90\le\omega\le 110$.
\begin{enumerate}
\item This model is arbitrage free.
\item There is an arbitrage involving buying two calls and selling one share of stock.
\item There is an arbitrage involving selling two calls and buying one share of stock.
\item There is an arbitrage involving buying one call and selling one share of stock.
\item There is an arbitrage involving selling one call and buying one share of stock.
\item none of the above.
\end{enumerate}
\end{xca}

\begin{xca}
As an option approaches expiration and the underlying tends to the strike,
\begin{enumerate}
\item the delta hedge may become unbounded.
\item the gamma hedge may oscillate between positive and negative values.
\item the option value tends to zero.
\item none of the above.
\item one of (a), (b), (c) is true.
\item two of (a), (b), (c) are true.
\item (a), (b), (c) are all true.
\end{enumerate}
\end{xca}

\begin{xca}
The dual cone of $\P = \{(p_1,p_2):0<p_1<p_2\}$ is
\begin{enumerate}
\item $\{(q_1,q_2):0<q_1<q_2\}$.
\item $\{(q_1,q_2):q_1<q_2\}$.
\item $\{(q_1,q_2):0<q_1 + q_2,q_2>0\}$.
\item $\{(q_1,q_2):0< q_1- q_2,q_2>0\}$.
\item none of the above.
\end{enumerate}
\end{xca}

\begin{xca}
Define the function $F$ that takes a matrix to its square, $F(X) = X^2$. The Fr\'echet derivative of $F$ is
\begin{enumerate}
\item $DF(X)H = 2XH$.
\item $DF(X)H = XH + HX$.
\item $DF(X)H = 2HX$.
\item all of the above.
\item none of the above.
\end{enumerate}
\end{xca}

\begin{xca}
Let $(B_t)_{t\ge0}$ be standard Brownian motion, then $\int_0^t B_s^2\,dB_s$ equals
\begin{enumerate}
\item $(B^3_t - t^{3/2})/3$.
\item $(B^3_t - t/2)/3$.
\item $(B^3_t - t^2)/3$.
\item $(B^3_t - t)/3$.
\item $(B^3_t - B_t)/3$.
\item $(B^3_t - tB_t)/3$.
\item none of the above.
\end{enumerate}

\end{xca}

\begin{xca}
Suppose a derivative contract is written against a risky zero coupon bond having 100 units notional, 
one year maturity, and no recovery 
that pays 1 unit if the bond is trading between 10 and 90 one day before maturity.
You may suppose the risk-free rate is 3\% and the hazard rate of default is 2\%.
The value of this derivative
\begin{enumerate}
\item is close to zero.
\item close to the one times the one year risk-free discount.
\item approximately 0.8 times the one year risk-free discount.
\item none of the above.
\item cannot be accurately determined from the given data.
\end{enumerate}
\end{xca}

\end{document}
